---
title: "Import data with R - Importing data from flat files with utils"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    css: [default, extra.css]
    includes:
      after_body: ../assets/html/footer.html
---

##Importing data from flat files with utils  

```{r, echo=TRUE, results='hide', message = FALSE}
# Load the knitr and kableExtra packages
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
# Load the gapminder package
library(gapminder)
# Load the dpylr package
library(dplyr)
# Load the ggplot2 package as well
library(ggplot2)
theme_set(theme_bw())  # pre-set the bw theme.
```


###Introduction & read.csv  

####read.csv  

The `utils` package, which is automatically loaded in your R session on startup, can import CSV files with the `read.csv()` function.  

In this exercise, you'll be working with `swimming_pools.csv`; it contains data on swimming pools in Brisbane, Australia (Source: data.gov.au). The file contains the column names in the first row. It uses a comma to separate values within rows.  

Type `dir()` in the console to list the files in your working directory. You'll see that it contains `swimming_pools.csv`, so you can start straight away.  

```{r}
dir()
path <- file.path("..", "xDatasets", "swimming_pools.csv")

# Import swimming_pools.csv: pools
pools <- read.csv(path)

# Print the structure of pools
str(pools)
```


####stringsAsFactors  

With `stringsAsFactors`, you can tell R whether it should convert strings in the flat file to factors.  

For all importing functions in the `utils` package, this argument is `TRUE`, which means that you import strings as factors. This only makes sense if the strings you import represent categorical variables in R. If you set `stringsAsFactors to FALSE`, the data frame columns corresponding to strings in your text file will be character.  

You'll again be working with the `swimming_pools.csv` file. It contains two columns (Name and Address), which shouldn't be factors.  

```{r}
path <- file.path("..", "xDatasets", "swimming_pools.csv")

# Import swimming_pools.csv correctly: pools
pools <- read.csv(path, stringsAsFactors = FALSE)

# Check the structure of pools
str(pools)
```


###read.delim & read.table  

####read.delim  

Aside from `.csv` files, there are also the `.txt` files which are basically text files. You can import these functions with `read.delim()`. By default, it sets the sep argument to `"\t"` (fields in a record are delimited by tabs) and the header argument to `TRUE` (the first row contains the field names).  

In this exercise, you will import `hotdogs.txt`, containing information on sodium and calorie levels in different hotdogs (Source: UCLA). The dataset has 3 variables, but the variable names are not available in the first line of the file. The file uses tabs as field separators. 

```{r}
path <- file.path("..", "xDatasets", "hotdogs.txt")

# Import hotdogs.txt: hotdogs
hotdogs <- read.delim(path, 
                      header = FALSE, 
                      col.names = c("type", "calories", "sodium"))

# Summarize hotdogs
sum_hotdogs <- as.data.frame(do.call(cbind, lapply(hotdogs[,2:3], summary)))

# Round to 2 decimals
sum_hotdogs <- round(sum_hotdogs,2)

sum_hotdogs %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```

####read.table  

If you're dealing with more exotic flat file formats, you'll want to use `read.table()`. It's the most basic importing function; you can specify tons of different arguments in this function. Unlike `read.csv()` and `read.delim()`, the header argument defaults to `FALSE` and the sep argument is `""` by default.  

Up to you again! The data is still `hotdogs.txt`. It has no column names in the first row, and the field separators are tabs. This time, though, the file is in the data folder inside your current working directory. A variable path with the location of this file is already coded for you.  

```{r}
# Path to the hotdogs.txt file: path
path <- file.path("..", "xDatasets", "hotdogs.txt")

# Import the hotdogs.txt file: hotdogs
hotdogs <- read.table(path, 
                      sep = "", 
                      col.names = c("type", "calories", "sodium"))

# Call head() on hotdogs
hotdogs %>%
  head %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


####Arguments  

Lily and Tom are having an argument because they want to share a hot dog but they can't seem to agree on which one to choose. After some time, they simply decide that they will have one each. Lily wants to have the one with the fewest calories while Tom wants to have the one with the most sodium.  

Next to calories and sodium, the hotdogs have one more variable: type. This can be one of three things: `Beef`, `Meat`, or `Poultry`, so a categorical variable: a factor is fine.  

```{r}
# Finish the read.delim() call
hotdogs <- read.delim(path, header = FALSE, col.names = c("type", "calories", "sodium"))

# Select the hot dog with the least calories: lily
lily <- hotdogs[which.min(hotdogs$calories), ]

# Select the observation with the most sodium: tom
tom <- hotdogs[which.max(hotdogs$sodium), ]

# Print lily and tom
lily %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

tom %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


####Column classes  

Next to `column names`, you can also specify the `column types` or `column classes` of the resulting data frame. You can do this by setting the colClasses argument to a vector of strings representing classes:  

```{r, eval=FALSE}
read.delim("my_file.txt", 
           colClasses = c("character",
                          "numeric",
                          "logical"))
```
                          
This approach can be useful if you have some columns that should be factors and others that should be characters. You don't have to bother with   `stringsAsFactors` anymore; just state for each column what the class should be.

If a column is set to `"NULL"` in the `colClasses` vector, this column will be skipped and will not be loaded into the data frame.  

```{r}
# Previous call to import hotdogs.txt
hotdogs <- read.delim(path, header = FALSE, col.names = c("type", "calories", "sodium"))

# Display structure of hotdogs
str(hotdogs)

# Edit the colClasses argument to import the data correctly: hotdogs2
hotdogs2 <- read.delim(path, header = FALSE, 
                       col.names = c("type", "calories", "sodium"),
                       colClasses = c("factor", "NULL", "numeric"))


# Display structure of hotdogs2
str(hotdogs2)
```

