---
title: "Import data with R - Importing data with readr & data.table"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    css: [default, extra.css]
    includes:
      after_body: ../assets/html/footer.html
---

##readr & data.table

```{r, echo=FALSE, results='hide', message = FALSE}
# Load the knitr and kableExtra packages
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
# Load the gapminder package
library(gapminder)
# Load the dpylr package
library(dplyr)
# Load the ggplot2 package as well
library(ggplot2)
theme_set(theme_bw())  # pre-set the bw theme.
```

###readr: read_csv & read_tsv  

####read_csv  

CSV files can be imported with `read_csv()`. It's a wrapper function around `read_delim()` that handles all the details for you. For example, it will assume that the first row contains the column names.  

The dataset you'll be working with here is `potatoes.csv`. It gives information on the impact of storage period and cooking on potatoes' flavor. It uses commas to delimit fields in a record, and contains column names in the first row. The file is available in your workspace. Remember that you can inspect your workspace with `dir()`.  

```{r, message = FALSE}
# Load the readr package
library(readr)

dir()
path <- file.path("..", "xDatasets", "potatoes.csv")

# Import potatoes.csv with read_csv(): potatoes
potatoes <- read_csv(path)
```


####read_tsv  

Where you use `read_csv()` to easily read in CSV files, you use `read_tsv()` to easily read in TSV files. `TSV is short for tab-separated values`.  

This time, the potatoes data comes in the form of a tab-separated values file; potatoes.txt is available in your workspace. In contrast to potatoes.csv, this file does not contain columns names in the first row, though.  

There's a vector properties that you can use to specify these column names manually.  

```{r}
# Load the readr package
library(readr)

path <- file.path("..", "xDatasets", "potatoes.txt")

# Column names
properties <- c("area", "temp", "size", "storage", "method",
                "texture", "flavor", "moistness")

# Import potatoes.txt: potatoes
potatoes <- read_tsv(path, col_names = properties)

# Call head() on potatoes
potatoes %>% 
  head %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


####read_delim  

Just as `read.table()` was the main utils function, `read_delim()` is the main `readr` function.   

`read_delim()` takes two mandatory arguments:  

> `file`: the file that contains the data  

> `delim`: the character that separates the values in the data file  

You'll again be working `potatoes.txt`; the file uses tabs `("\t")` to delimit values and does not contain column names in its first line. It's available in your working directory so you can start right away. As before, the vector properties is available to set the col_names.  

```{r}
# readr is already loaded
library(readr)

# Column names
properties <- c("area", "temp", "size", "storage", "method",
                "texture", "flavor", "moistness")

# Import potatoes.txt using read_delim(): potatoes
potatoes <- read_delim(path, delim = "\t", col_names = properties)

# Print out potatoes
potatoes %>% 
  tail %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


####skip and n_max  

Through `skip` and `n_max` you can control which part of your flat file you're actually importing into R.  

> `skip` specifies the number of lines you're ignoring in the flat file before actually starting to import data.  

> `n_max` specifies the number of lines you're actually importing.  

Say for example you have a CSV file with 20 lines, and set `skip = 2` and `n_max = 3`, you're only reading in lines 3, 4 and 5 of the file.  

Watch out: Once you skip some lines, you also skip the first line that can contain column names!  

```{r}
# Column names
properties <- c("area", "temp", "size", "storage", "method",
                "texture", "flavor", "moistness")

# Import 5 observations from potatoes.txt: potatoes_fragment
potatoes_fragment <- read_tsv(path, skip = 6, n_max = 5, col_names = properties)

potatoes_fragment %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


####col_types  

You can also specify which types the columns in your imported data frame should have. You can do this with `col_types`. If set to `NULL`, the default, functions from the readr package will try to find the correct types themselves. You can manually set the types with a string, where each character denotes the class of the column: `character, double, integer and logical`. `_` skips the column as a whole.  

```{r}
# Column names
properties <- c("area", "temp", "size", "storage", "method",
                "texture", "flavor", "moistness")

# Import all data, but force all columns to be character: potatoes_char
potatoes_char <- read_tsv(path, col_types = "cccccccc", col_names = properties)

# Print out structure of potatoes_char
str(potatoes_char)
```


####col_types with collectors  

Another way of setting the types of the imported columns is using `collectors`. `Collector functions`` can be passed in a `list()` to the `col_types` argument of read_ functions to tell them how to interpret values in a column.  

For a complete list of collector functions, you can take a look at the collector documentation. For this exercise you will need two collector functions:  

> `col_integer()`: the column should be interpreted as an integer.  

> `col_factor(levels, ordered = FALSE)`: the column should be interpreted as a factor with levels.  

```{r}
path <- file.path("..", "xDatasets", "hotdogs.txt")

# Import without col_types
hotdogs <- read_tsv(path, col_names = c("type", "calories", "sodium"))

# Display the summary of hotdogs
summary(hotdogs)

# The collectors you will need to import the data
fac <- col_factor(levels = c("Beef", "Meat", "Poultry"))
int <- col_integer()

# Edit the col_types argument to import the data correctly: hotdogs_factor
hotdogs_factor <- read_tsv("../xDatasets/hotdogs.txt",
                           col_names = c("type", "calories", "sodium"),
                           col_types = list(fac, int, int))

# Display the summary of hotdogs_factor
do.call(cbind, lapply(hotdogs_factor, summary))
```



###data.table: fread  

####fread  

You still remember how to use `read.table()`, right? Well, `fread()` is a function that does the same job with very similar arguments. It is extremely easy to use and blazingly fast! Often, simply specifying the path to the file is enough to successfully import your data.  

Don't take our word for it, try it yourself! You'll be working with the `potatoes.csv` file, that's available in your workspace. Fields are delimited by commas, and the first line contains the column names. 

```{r, echo=FALSE, results='hide', message = FALSE, eval = FALSE}
install.packages("data.table")
```


```{r, message = FALSE}
# load the data.table package
library(data.table)

path <- file.path("..", "xDatasets", "potatoes.csv")

# Import potatoes.csv with fread(): potatoes
potatoes <- fread(path)

# Print out potatoes
potatoes %>% 
  tail() %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


####fread: more advanced use  

Now that you know the basics about `fread()`, you should know about two arguments of the function: `drop` and `select`, to drop or select variables of interest.  

Suppose you have a dataset that contains 5 variables and you want to keep the first and fifth variable, named "a" and "e". The following options will all do the trick:  

> `fread("path/to/file.txt", drop = 2:4)`    
`fread("path/to/file.txt", select = c(1, 5))`    
`fread("path/to/file.txt", drop = c("b", "c", "d"))`    
`fread("path/to/file.txt", select = c("a", "e"))`    

Let's stick with potatoes since we're particularly fond of them here at DataCamp. The data is again available in the file potatoes.csv, containing comma-separated records.  

```{r}
# Import columns 6 and 8 of potatoes.csv: potatoes
potatoes <- fread("../xDatasets/potatoes.csv", select = c(6, 8))

# Plot texture (x) and moistness (y) of potatoes
plot(x = potatoes$texture, y = potatoes$moistness)
```

