---
title: "Import data with R - Importing Excel Data"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    css: [default, extra.css]
    includes:
      after_body: ../assets/html/footer.html
---

```{r, echo=FALSE, results='hide', message = FALSE}
# Load the knitr and kableExtra packages
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
# Load the gapminder package
library(gapminder)
# Load the dpylr package
library(dplyr)
# Load the ggplot2 package as well
library(ggplot2)
theme_set(theme_bw())  # pre-set the bw theme.
```

##readxl package  

####List the sheets of an Excel file  

Before you can start importing from Excel, you should find out which sheets are available in the workbook. You can use the `excel_sheets()` function for this.  

You will find the Excel file `urbanpop.xlsx` in your working directory (type `dir()` to see it). This dataset contains urban population metrics for practically all countries in the world throughout time (Source: Gapminder). It contains three sheets for three different time periods. In each sheet, the first row contains the column names.  

```{r, echo=FALSE, results='hide', message = FALSE, eval = FALSE}
install.packages("readxl")
```


```{r}
# Load the readxl package
library(readxl)

# Print the names of all worksheets
excel_sheets("../xDatasets/urbanpop.xlsx")
```

*As you can see, the result of `excel_sheets()` is simply a character vector; you haven't imported anything yet. That's something for the `read_excel()` function.*  


####Import an Excel sheet  

Now that you know the names of the sheets in the Excel file you want to import, it is time to import those sheets into R. You can do this with the `read_excel()` function. Have a look at this recipe:  

> `data <- read_excel("data.xlsx", sheet = "my_sheet")`  

This call simply imports the sheet with the name `"my_sheet"` from the `"data.xlsx"` file. You can also pass a number to the sheet argument; this will cause `read_excel()` to import the sheet with the given sheet number. `sheet = 1` will import the first sheet, `sheet = 2` will import the second sheet, and so on.  

```{r}
# Read the sheets, one by one
pop_1 <- read_excel("../xDatasets/urbanpop.xlsx", sheet = 1)
pop_2 <- read_excel("../xDatasets/urbanpop.xlsx", sheet = 2)
pop_3 <- read_excel("../xDatasets/urbanpop.xlsx", sheet = 3)

# Put pop_1, pop_2 and pop_3 in a list: pop_list
pop_list = list(pop_1, pop_2, pop_3)

# Display the structure of pop_list
str(pop_list)
```

*we will learn how to use both the `excel_sheets()` and the `read_excel()` function in combination with `lapply()` to read multiple sheets at once.*  


####Reading a workbook  

In the previous exercise you generated a list of three Excel sheets that you imported. However, loading in every sheet manually and then merging them in a list can be quite tedious. Luckily, you can automate this with `lapply()`.  

Have a look at the example code below:  

```{r, eval=FALSE}
my_workbook <- lapply(excel_sheets("data.xlsx"),
                      read_excel,
                      path = "data.xlsx")
```

The `read_excel()` function is called multiple times on the `"data.xlsx"` file and each sheet is loaded in one after the other. The result is a list of data frames, each data frame representing one of the sheets in `data.xlsx`.  

```{r}
# Read all Excel sheets with lapply(): pop_list
pop_list <- lapply(excel_sheets("../xDatasets/urbanpop.xlsx"),
                   read_excel,
                   path = "../xDatasets/urbanpop.xlsx")

# Display the structure of pop_list
str(pop_list)
```


####The col_names argument  

Apart from `path` and `sheet`, there are several other arguments you can specify in `read_excel()`. One of these arguments is called `col_names`.  

By default it is `TRUE`, denoting whether the first row in the Excel sheets contains the column names. If this is not the case, you can set col_names to FALSE. In this case, R will choose column names for you. You can also choose to set col_names to a character vector with names for each column. It works exactly the same as in the readr package.  

You'll be working with the `urbanpop_nonames.xlsx` file. It contains the same data as `urbanpop.xlsx` but has `no column names in the first row` of the excel sheets.  

```{r}
# Import the first Excel sheet of urbanpop_nonames.xlsx (R gives names): pop_a
pop_a <- read_excel("../xDatasets/urbanpop_nonames.xlsx", col_names = FALSE)

# Import the first Excel sheet of urbanpop_nonames.xlsx (specify col_names): pop_b
cols <- c("country", paste0("year_", 1960:1966))
pop_b <- read_excel("../xDatasets/urbanpop_nonames.xlsx", col_names = cols)

# Print the summary of pop_a
sum_pop_a <- as.data.frame(do.call(cbind, lapply(pop_a, summary)))
sum_pop_a %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

# Print the summary of pop_b
sum_pop_b <- as.data.frame(do.call(cbind, lapply(pop_b, summary))) 
sum_pop_b %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```

*Did you spot the difference between the summaries? It's really crucial to correctly tell R whether your Excel data contains column names. If you don't, the head of the data frame you end up with will contain incorrect information.*  


####The skip argument  

Another argument that can be very useful when reading in Excel files that are less tidy, is `skip`. With `skip`, you can tell R to ignore a specified number of rows inside the Excel sheets you're trying to pull data from. Have a look at this example:  

> `read_excel("data.xlsx", skip = 15)`  

In this case, the `first 15 rows in the first sheet of "data.xlsx" are ignored`.  

If the first row of this sheet contained the column names, this information will also be ignored by readxl. Make sure to set `col_names` to `FALSE` or manually specify column names in this case!  

```{r}
# Import the second sheet of urbanpop.xlsx, skipping the first 21 rows: urbanpop_sel
urbanpop_sel <- read_excel("../xDatasets/urbanpop.xlsx", sheet = 2, col_names = FALSE, skip = 21)

# Print out the first observation from urbanpop_sel
urbanpop_sel[1,]
```

*Time to learn about another package to import data from Excel: gdata.*  

##gdata package     

```{r, echo=FALSE, results='hide', message = FALSE, eval = FALSE}
install.packages("gdata")
```

####Import a local file  

In this part of the chapter you'll learn how to import `.xls` files using the `gdata` package. Similar to the `readxl` package, you can import single Excel sheets from Excel sheets to start your analysis in R.  

You'll be working with the `urbanpop.xls` dataset, the `.xls` version of the Excel file you've been working with before.  

```{r, message = FALSE}
# Load the gdata package
library(gdata)

prl <- "C:/myperl/perl/bin/perl5.28.1.exe"

# Import the second sheet of urbanpop.xls: urban_pop
urban_pop <- read.xls("../xDatasets/urbanpop.xls", perl = prl)

# Print the first 11 observations using head()
urban_pop %>% 
  head(11) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```

*There seems to be a lot of missing data, but `read.xls()` knows how to handle it.*  


####read.xls() wraps around read.table()  

Remember how `read.xls()` actually works? It basically comes down to two steps: converting the Excel file to a `.csv` file using a `Perl script`, and then reading that `.csv` file with the `read.csv()` function that is loaded by default in R, through the `utils` package.  

This means that all the options that you can specify in `read.csv()`, can also be specified in `read.xls()`.  

```{r, message = FALSE}
# Column names for urban_pop
columns <- c("country", paste0("year_", 1967:1974))

# Finish the read.xls call
urban_pop <- read.xls("../xDatasets/urbanpop.xls", sheet = 2,
                      skip = 50, header = FALSE, stringsAsFactors = FALSE,
                      col.names = columns,
                     perl = prl)

# Print first 10 observation of urban_pop
urban_pop[1:10, ] %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

```



####Work that Excel data!  

Now that you can read in Excel data, let's try to clean and merge it. You already used the `cbind()` function some exercises ago. Let's take it one step further now.  
Make sure the first column of `urban_sheet2` and `urban_sheet3` are removed, so you don't have duplicate columns.  

```{r, message = FALSE}
# Add code to import data from all three sheets in urbanpop.xls
path <- "../xDatasets/urbanpop.xls"
urban_sheet1 <- read.xls(path, sheet = 1, stringsAsFactors = FALSE,  perl = prl)
urban_sheet2 <- read.xls(path, sheet = 2, stringsAsFactors = FALSE,  perl = prl)
urban_sheet3 <- read.xls(path, sheet = 3, stringsAsFactors = FALSE,  perl = prl)

# Extend the cbind() call to include urban_sheet3: urban
urban <- cbind(urban_sheet1, urban_sheet2[-1], urban_sheet3[-1])

# Remove all rows with NAs from urban: urban_clean
urban_clean <- na.omit(urban)

# Print out a summary of urban_clean
sum_urban_clean <- as.data.frame(do.call(cbind, lapply(urban_clean, summary)))

sum_urban_clean %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


**Time for something totally different: XLConnect**

