---
title: "Data Manipulation in R with dplyr - Select and mutate"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    css: [default, extra.css]
    includes:
      after_body: ../assets/html/footer.html
---


```{r, echo=FALSE, results='hide', message = FALSE}
# Load the knitr and kableExtra packages
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
# Load the gapminder package
library(gapminder)
# Load the dpylr package
library(dplyr)
# Load the ggplot2 package as well
library(ggplot2)
theme_set(theme_bw())  # pre-set the bw theme.
```

##Section 3 - The five verbs and select in more detail  

###The five verbs and their meaning  

The `dplyr` package contains **five** key data manipulation functions, also called verbs:  

> `select()`, which returns a subset of the columns,  
> `filter()`, that is able to return a subset of the rows,  
> `arrange()`, that reorders the rows according to single or multiple variables,
> `mutate()`, used to add columns from existing data,  
> `summarize()`, which reduces each group to a single row by calculating aggregate measures.  

If you want to find out more about these functions, consult the documentation by clicking on the functions above.  



###Choosing is not losing! The select verb  

To answer the simple question whether flight delays tend to shrink or grow during a flight, we can safely discard a lot of the variables of each flight. To select only the ones that matter, we can use `select()`.  

As an example, take the following call, that selects the variables `var1` and `var2` from the data frame `df`.

> `select(df, var1, var2)`  

You can also use : to select a range of variables and `-` to exclude some variables, similar to indexing a data.frame with square brackets. You can use both variable's names as well as integer indexes. This call selects the four first variables except for the second one of a data frame df:

> `select(df, 1:4, -2)`  

select() does not change the data frame it is called on; you have to explicitly assign the result of `select()` to a variable to store the result.  

```{r}
# Load the hflights package
library(hflights)

hflights_df <- hflights[sample(nrow(hflights), 720), ] 
hflights <- as_tibble(hflights_df)

# Print out a tbl with the four columns of hflights related to delay
head(select(hflights, ActualElapsedTime, AirTime, ArrDelay, DepDelay))

# Print out the columns Origin up to Cancelled of hflights
head(select(hflights, 14:19))

# Answer to last question: be concise!
head(select(hflights, 1:4, 12:21))
```

*Bear in mind that select() does not modify the original dataset. Instead, select() returns a new, modified copy. Therefore, choosing is not losing! You should save the result to a new variable if you want to be able to use it later.*  



###Helper functions for variable selection  

`dplyr` comes with a set of helper functions that can help you select groups of variables inside a `select()` call:  

```{r eval = FALSE}
starts_with("X")  # every name that starts with "X",
ends_with("X")    # every name that ends with "X",
contains("X")     # every name that contains "X",
matches("X")      # every name that matches "X", where "X" can be a regular expression,
num_range("x", 1:5) # the variables named x01, x02, x03, x04 and x05,
one_of(x)         # every name that appears in x, which should be a character vector.
```


Pay attention here: When you refer to columns directly inside select(), you don't use quotes. If you use the helper functions, you do use quotes.

```{r}
# Print out a tbl containing just ArrDelay and DepDelay
head(select(hflights, ends_with(c('Delay'))))

# Print out a tbl as described in the second instruction, using both helper functions and variable names
head(select(hflights, UniqueCarrier, ends_with(c('Num')), starts_with(c('Cancel'))))

# Print out a tbl as described in the third instruction, using only helper functions.
head(select(hflights, ends_with("Time"), ends_with("Delay")))
```

*This exercise demonstrates the power of the helper functions. They can limit the amount of coding drastically. Beware of too obscure queries however; you still want your code to be clear for fellow data scientists!*


###Comparison to base R  

To see the added value of the `dplyr` package, it is useful to compare its syntax with base R. Up to now, you have only considered functionality that is also available without the use of `dplyr`. The elegance and ease-of-use of `dplyr` is a great plus though.  

```{r}
# Finish select call so that ex1d matches ex1r
ex1r <- hflights[c("TaxiIn", "TaxiOut", "Distance")]
ex1d <- select(hflights, starts_with("Taxi"), Distance)

# Finish select call so that ex2d matches ex2r
ex2r <- hflights[c("Year", "Month", "DayOfWeek", "DepTime", "ArrTime")]
ex2d <- select(hflights, Year, Month, DayOfWeek, DepTime, ArrTime)

# Finish select call so that ex3d matches ex3r
ex3r <- hflights[c("TailNum", "TaxiIn", "TaxiOut")]
ex3d <- select(hflights, TailNum, starts_with("Taxi"))
```



##Section 4 - The second of five verbs: mutate  

###Mutating is creating  

`mutate()` is the second of five data manipulation functions you will get familiar with in this course. `mutate()` creates new columns which are added to a copy of the dataset.  

Take this example that adds a new column, `z`, which is the element-wise sum of the columns `x` and `y`, to the data frame `df`:  

> `mutate(df, z = x + y)`  

```{r}
# Add the new variable ActualGroundTime to a copy of hflights and save the result as g1.
g1 <- mutate(hflights, ActualGroundTime = ActualElapsedTime - AirTime)

# Add the new variable GroundTime to g1. Save the result as g2.
g2 <- mutate(g1, GroundTime = TaxiIn + TaxiOut)

# Add the new variable AverageSpeed to g2. Save the result as g3.
g3 <- mutate(g2, AverageSpeed = 60 * Distance / AirTime)

# Print out g3
g3 %>%
  head(10) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "center", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```



###Add multiple variables using mutate  

So far you've added variables to `hflights` one at a time, but you can also use `mutate()` to add multiple variables at once. To create more than one variable, place a comma between each variable that you define inside `mutate()`.  

`mutate()` even allows you to use a new variable while creating a next variable in the same call. In this example, the new variable `x` is directly reused to create the new variable `y`:  

> `mutate(my_df, x = a + b, y = x + c)`  


```{r}
# Add a second variable loss_ratio to the dataset: m1
m1 <- mutate(hflights, loss = ArrDelay - DepDelay, loss_ratio = loss / DepDelay)

# Add the three variables as described in the third instruction: m2
m2 <- mutate(hflights, TotalTaxi = TaxiIn + TaxiOut, ActualGroundTime = ActualElapsedTime - AirTime, Diff = TotalTaxi - ActualGroundTime)
```


###Session info

```{r}
sessionInfo()
```
