---
title: "Data Manipulation in R with dplyr - Filter and arrange"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    css: [default, extra.css]
    includes:
      after_body: ../assets/html/footer.html
---


```{r, echo=FALSE, results='hide', message = FALSE}
# Load the knitr and kableExtra packages
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
# Load the gapminder package
library(gapminder)
# Load the dpylr package
library(dplyr)
# Load the ggplot2 package as well
library(ggplot2)
theme_set(theme_bw())  # pre-set the bw theme.
```

##Section 5 - The third of five verbs: filter  

###Logical operators  

R comes with a set of logical operators that you can use inside `filter()`:  

> `x < y`, `TRUE` if `x` is less than `y`  
> `x <= y`, TRUE` if `x` is less than or equal to `y`  
> `x == y`, TRUE` if `x` equals `y`  
> `x != y`, TRUE` if `x` does not equal `y`  
> `x >= y`, TRUE` if `x` is greater than or equal to `y`  
> `x > y`, TRUE` if `x` is greater than `y`  
> `x %in% c(a, b, c)`, `TRUE` if `x` is in the vector `c(a, b, c)`  

The following example filters df such that only the observations for which a is positive, are kept:  

> `filter(df, a > 0)`  

```{r}
# Load the hflights package
library(hflights)

hflights_df <- hflights[sample(nrow(hflights), 720), ] 
hflights <- as_tibble(hflights)

# All flights that traveled 3000 miles or more
filter(hflights, Distance >= 3000) %>% glimpse()

# All flights flown by one of JetBlue, Southwest, or Delta
filter(hflights, UniqueCarrier %in% c("JetBlue", "Southwest", "Delta")) %>% glimpse()

# All flights where taxiing took longer than flying
filter(hflights, (TaxiIn + TaxiOut) > AirTime) %>% glimpse()
```



###Combining tests using boolean operators  

R also comes with a set of boolean operators that you can use to combine multiple logical tests into a single test. These include `&` (and), `|` (or), and `!` (not). Instead of using the `&` operator, you can also pass several logical tests to `filter()`, separated by commas. The following two calls are completely equivalent:  

> `filter(df, a > 0 & b > 0)`  
> `filter(df, a > 0, b > 0)`  

Next, `is.na()` will also come in handy. This example keeps the observations in df for which the variable `x` is not `NA`:  

> `filter(df, !is.na(x))`  

```{r}
# All flights that departed before 5am or arrived after 10pm
filter(hflights, DepTime < 500 | ArrTime > 2200) %>% glimpse()

# All flights that departed late but arrived ahead of schedule
filter(hflights, DepDelay > 0, ArrDelay < 0) %>% glimpse()

# All flights that were cancelled after being delayed
filter(hflights, DepDelay > 0, Cancelled == 1) %>% glimpse()
```



###Blend together what you've learned!  

So far, you have learned three data manipulation functions in the `dplyr` package. Time for a summarizing exercise. You will generate a new dataset from the ``hflights dataset that contains some useful information on flights that had `JFK airport` as their destination. You will need `select()`, `mutate()` and `filter()`.  

```{r}
# Select the flights that had JFK as their destination: c1
c1 <- filter(hflights, Dest == "JFK")

# Combine the Year, Month and DayofMonth variables to create a Date column: c2
c2 <- mutate(c1, Date = paste(Year, Month, DayofMonth, sep='-'))

# Print out a selection of columns of c2
select(c2, Date, DepTime, ArrTime, TailNum)
```



###Recap on select, mutate and filter  

With `select()`, `mutate()` and `filter()`, you can already reveal interesting information from a dataset. Through a combination of these expressions or by the use of a one-liner, try to answer the following question:  

*How many weekend flights flew a distance of more than 1000 miles but had a total taxiing time below 15 minutes?*

```{r}
filter(hflights, DayOfWeek %in% c(6,7), Distance > 1000, (TaxiIn + TaxiOut) < 15) %>% glimpse()
```

answer: 155 flights !


##Section 6 - Almost there: the arrange verb  

###Arranging your data  

`arrange()` can be used to rearrange rows according to any type of data. If you pass `arrange()` a character variable, for example, R will rearrange the rows in alphabetical order according to values of the variable. If you pass a factor variable, R will rearrange the rows according to the order of the levels in your factor (running levels() on the variable reveals this order).  

`dtc` has already been defined on the right. It's up to you to write some `arrange()` expressions to display its contents appropriately!  

```{r}
# Definition of dtc
dtc <- filter(hflights, Cancelled == 1, !is.na(DepDelay))

# Arrange dtc by departure delays
arrange(dtc, DepDelay) %>%
  head(10) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "center", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

# Arrange dtc so that cancellation reasons are grouped
arrange(dtc, CancellationCode) %>%
  head(10) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "center", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

# Arrange dtc according to carrier and departure delays
arrange(dtc, UniqueCarrier, DepDelay) %>%
  head(10) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "center", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```



###Reverse the order of arranging  

By default, `arrange()` arranges the rows from `smallest` to `largest`. Rows with the smallest value of the variable will appear at the top of the data set. You can reverse this behavior with the `desc()` function. `arrange()` will reorder the rows from `largest` to `smallest` values of a variable if you wrap the variable name in `desc()` before passing it to `arrange()`.  
```{r}
# Arrange according to carrier and decreasing departure delays
arrange(hflights, UniqueCarrier, desc(DepDelay))  %>%
  head(10) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "center", font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")


# Arrange flights by total delay (normal order).
arrange(hflights, (DepDelay + ArrDelay))  %>%
  head(10) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "center", font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```



###Session info

```{r}
sessionInfo()
```
