---
title: "Intermediate of R - Loops"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
## While Loops  

####Write a while loop  


Let's get you started with building a while loop from the ground up. Have another look at its recipe:  

```{r, eval=FALSE}
while (condition) { 
  expr
}
```

Remember that the `condition` part of this recipe should become FALSE at some point during the execution. Otherwise, the `while` loop will go on indefinitely. In DataCamp's learning interface, your session will be disconnected in this case.  

Have a look at the code on the right; it initializes the `speed` variables and already provides a `while loop` template to get you started.  

```{r}
# Initialize the speed variable
speed <- 64

# Code the while loop
while (speed > 30) {
  print("Slow down!")
  speed <- speed - 7
}

# Print out the speed variable
speed
```

 
####Throw in more conditionals  

In the previous exercise, you simulated the interaction between a driver and a driver's assistant: When the speed was too high, "Slow down!" got printed out to the console, resulting in a decrease of your speed by 7 units.  

There are several ways in which you could make your driver's assistant more advanced. For example, the assistant could give you different messages based on your speed or provide you with a current speed at a given moment.  

A while loop similar to the one you've coded in the previous exercise is already available in the editor. It prints out your current speed, but there's no code that decreases the `speed` variable yet, which is pretty dangerous. Can you make the appropriate changes?  

```{r}
# Initialize the speed variable
speed <- 64

# Extend/adapt the while loop
while (speed > 30) {
  print(paste("Your speed is",speed))
  if (speed > 48) {
    print("Slow down big time!")
    speed <- speed - 11
  } else {
    print("Slow down!")
    speed <- speed - 6
  }
}
```

####Stop the while loop: break  

There are some very rare situations in which severe speeding is necessary: what if a hurricane is approaching and you have to get away as quickly as possible? You don't want the driver's assistant sending you speeding notifications in that scenario, right?  

This seems like a great opportunity to include the `break` statement in the while loop you've been working on. Remember that the `break` statement is a control statement. When R encounters it, the while loop is abandoned completely.  

```{r}
# Initialize the speed variable
speed <- 88

while (speed > 30) {
  print(paste("Your speed is", speed))
  
  # Break the while loop when speed exceeds 80
  if (speed > 80) {
    break
  }
  
  if (speed > 48) {
    print("Slow down big time!")
    speed <- speed - 11
  } else {
    print("Slow down!")
    speed <- speed - 6
  }
}
```


####Build a while loop from scratch  

The previous exercises guided you through developing a pretty advanced while loop, containing a break statement and different messages and updates as determined by control flow constructs. If you manage to solve this comprehensive exercise using a while loop, you're totally ready for the next topic: the for loop.  

```{r}
# Initialize i as 1 
i <- 1

# Code the while loop
while (i <= 10) {
  print(3 * i)
  if ((3 * i) %% 8 == 0 ) {
    print(3 * i)
    break
  }
  i <- i + 1
}
```




## For Loops  

####Loop over a vector  

In the previous video, Filip told you about two different strategies for using the for loop. To refresh your memory, consider the following loops that are equivalent in R:  

```{r}
primes <- c(2, 3, 5, 7, 11, 13)

# loop version 1
for (p in primes) {
  print(p)
}

# loop version 2
for (i in 1:length(primes)) {
  print(primes[i])
}
```

Remember our `linkedin` vector? It's a vector that contains the number of views your LinkedIn profile had in the last seven days. The `linkedin` vector has already been defined in the editor on the right so that you can fully focus on the instructions!  


```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Loop version 1
for (l in linkedin){
  print(l)
}

# Loop version 2
for (i in 1:length(linkedin)){
  print(linkedin[i])
}
```


####Loop over a list  

Looping over a list is just as easy and convenient as looping over a vector. There are again two different approaches here:  

```{r}
primes_list <- list(2, 3, 5, 7, 11, 13)

# loop version 1
for (p in primes_list) {
  print(p)
}

# loop version 2
for (i in 1:length(primes_list)) {
  print(primes_list[[i]])
}
```


Notice that you need double square brackets - `[[ ]]` - to select the list elements in loop version 2.  

Suppose you have a list of all sorts of information on New York City: its population size, the names of the boroughs, and whether it is the capital of the United States. We've already prepared a list `nyc` with all this information in the editor (source: Wikipedia).  

```{r}
# The nyc list is already specified
nyc <- list(pop = 8405837, 
            boroughs = c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island"), 
            capital = FALSE)

# Loop version 1
for (n in nyc){
  print(n)
}

# Loop version 2
for (i in 1:length(nyc)){
  print(nyc[[i]])
}
```


####Loop over a matrix  

In your workspace, there's a matrix `ttt`, that represents the status of a tic-tac-toe game. It contains the values "X", "O" and "NA". Print out `ttt` in the console so you can have a closer look. On row 1 and column 1, there's "O", while on row 3 and column 2 there's "NA".  

To solve this exercise, you'll need a `for` loop inside a `for` loop, often called a nested loop. Doing this in R is a breeze! Simply use the following recipe:  

```{r, eval=FALSE}
for (var1 in seq1) {
  for (var2 in seq2) {
    expr
  }
}
```


```{r}
ttt = matrix( 
  c("0", NA, "X", NA, "0", NA, "X", "0", "X"),
  nrow=3, 
  ncol=3)

# The tic-tac-toe matrix ttt
ttt
```


```{r}
# define the double for loop
for (i in 1:nrow(ttt)) {
  for (j in 1:ncol(ttt)) {
    print(paste("On row", i ,"and column", j ,"the board contains ", ttt[[i,j]]))
  }
}
```


####Mix it up with control flow  

Let's return to the *LinkedIn* profile views data, stored in a vector linkedin. In the first exercise on `for` loops you already did a simple printout of each element in this vector. A little more in-depth interpretation of this data wouldn't hurt, right? Time to throw in some conditionals! As with the `while` loop, you can use the `if` and `else` statements inside the for loop.  

```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Code the for loop with conditionals
for (li in linkedin) {
  if (li > 10) {
    print("You're popular!")
  } else {
    print("Be more visible!")
  }
  print(li)
}
```


####Next, you break it  

In the editor on the right you'll find a possible solution to the previous exercise. The code loops over the `linkedin` vector and prints out different messages depending on the values of li.  

In this exercise, you will use the `break` and `next` statements:  

> The `break` statement abandons the active loop: the remaining code in the loop is skipped and the loop is not iterated over anymore.  

> The `next` statement skips the remainder of the code in the loop, but continues the iteration.  

```{r}
# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Adapt/extend the for loop
for (li in linkedin) {
  if (li > 10) {
    print("You're popular!")
  } else {
    print("Be more visible!")
  }
  
  # Add if statement with break
  if (li > 16){
    print("This is ridiculous, I'm outta here!")
    break
  }
  
  # Add if statement with next
  if (li < 5){
    print("This is too embarrassing!")
    next
  }
  
  print(li)
}
```


####Build a for loop from scratch  

This exercise will not introduce any new concepts on for loops.  

In the editor on the right, we already went ahead and defined a variable `rquote`. This variable has been split up into a vector that contains separate letters and has been stored in a vector `chars` with the `strsplit()` function.  

Can you write code that counts the number of r's that come before the first u in `rquote`?  

```{r}
# Pre-defined variables
rquote <- "r's internals are irrefutably intriguing"
chars <- strsplit(rquote, split = "")[[1]]

# Initialize rcount
rcount <- 0

# Finish the for loop
for (char in chars) {
  if (char == "r"){
    rcount <- rcount +1
  }
  if (char == "u"){
    break
  }
  print(rcount)
}

```

