---
title: "Functions - LAPPLY, SAPPLY & VAPPLY"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

##lapply

####Use lapply with a built-in R function  

Before you go about solving the exercises below, have a look at the documentation of the `lapply()` function. The Usage section shows the following expression:  

>`lapply(X, FUN, ...)`  

To put it generally, lapply takes a vector or list X, and applies the function FUN to each of its members. If FUN requires additional arguments, you pass them after you've specified X and FUN (...). The output of lapply() is a list, the same length as X, where each element is the result of applying FUN on the corresponding element of X.  

Now that you are truly brushing up on your data science skills, let's revisit some of the most relevant figures in data science history. We've compiled a vector of famous mathematicians/statisticians and the year they were born. Up to you to extract some information!  

```{r}
# The vector pioneers has already been created for you
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")

# Split names from birth year
split_math <- strsplit(pioneers, split = ":")

# Convert to lowercase strings: split_low
split_low <- lapply(split_math, tolower)

# Take a look at the structure of split_low
str(split_low)
```


####Use lapply with your own function  

As Filip explained in the instructional video, you can use `lapply()` on your own functions as well. You just need to code a new function and make sure it is available in the workspace. After that, you can use the function inside `lapply()` just as you did with base R functions.  

In the previous exercise you already used `lapply()` once to convert the information about your favorite pioneering statisticians to a list of vectors composed of two character strings. Let's write some code to select the names and the birth years separately.  

The sample code already includes code that defined `select_first()`, that takes a vector as input and returns the first element of this vector.  

```{r}
# Code from previous exercise:
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
split <- strsplit(pioneers, split = ":")
split_low <- lapply(split, tolower)

# Write function select_first()
select_first <- function(x) {
  x[1]
}

# Apply select_first() over split_low: names
names <- lapply(split_low, select_first)
unlist(names)

# Write function select_second()
select_second <- function(x) {
  x[2]
}

# Apply select_second() over split_low: years
years <- lapply(split_low, select_second)
unlist(years)
```


####lapply and anonymous functions  

Writing your own functions and then using them inside `lapply()` is quite an accomplishment! But defining functions to use them only once is kind of overkill, isn't it? That's why you can use so-called anonymous functions in R.  

Previously, you learned that functions in R are objects in their own right. This means that they aren't automatically bound to a name. When you create a function, you can use the assignment operator to give the function a name. It's perfectly possible, however, to not give the function a name. This is called an anonymous function:  

```{r}
# Named function
triple <- function(x) { 3 * x }

# Anonymous function with same implementation
function(x) { 3 * x }

# Use anonymous function inside lapply()
lapply(list(1,2,3), function(x) { 3 * x })
```

```{r}
# split_low has been created for you
split_low

# Transform: use anonymous function inside lapply
names <- lapply(split_low, function(x) {x[1]})

# Transform: use anonymous function inside lapply
years <- lapply(split_low, function(x) {x[2]})
```

Now, there's another way to solve the issue of using the `select_*()` functions only once: you can make a more generic function that can be used in more places.  


####Use lapply with additional arguments  

In the video, the `triple()` function was transformed to the `multiply()` function to allow for a more generic approach. `lapply()` provides a way to handle functions that require more than one argument, such as the `multiply()` function:

```{r}
multiply <- function(x, factor) {
  x * factor
}
lapply(list(1,2,3), multiply, factor = 3)
```

On the right we've included a generic version of the select functions that you've coded earlier: `select_el()`. It takes a vector as its first argument, and an index as its second argument. It returns the vector's element at the specified index.  

```{r}
# Definition of split_low
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
split <- strsplit(pioneers, split = ":")
split_low <- lapply(split, tolower)

# Generic select function
select_el <- function(x, index) {
  x[index]
}

# Use lapply() twice on split_low: names and years
names <- lapply(split_low, select_el, index=1)
years <- lapply(split_low, select_el, index=2)
```

*Did you notice that lapply() always returns a list, no matter the input? This can be kind of annoying. In the next video tutorial you'll learn about sapply() to solve this!!!*



##sapply  

####How to use sapply  

You can use `sapply()` similar to how you used `lapply()`. The first argument of `sapply()` is the list or vector X over which you want to apply a function, FUN. Potential additional arguments to this function are specified afterwards `(...)`:  

>`sapply(X, FUN, ...)`  

In the next couple of exercises, you'll be working with the variable `temp`, that contains temperature measurements for 7 days. `temp` is a list of length 7, where each element is a vector of length 5, representing 5 measurements on a given day. This variable has already been defined in the workspace: type `str(temp)` to see its structure.  


```{r}
# temp list of vectors
temp <- list(c(3,7,9,6,-1),c(6,9,12,13,5), c(4,8,3,-1,-3), c(1,4,7,2,-2), c(5,7,9,4,2), c(3,2,5,-2,1), c(3,1,12,-5,5))

# temp has already been defined in the workspace
str(temp)

# Use lapply() to find each day's minimum temperature
lapply(temp, min)

# Use sapply() to find each day's minimum temperature
sapply(temp, min)

# Use lapply() to find each day's maximum temperature
lapply(temp, max)

# Use sapply() to find each day's maximum temperature
sapply(temp, max)
```


*Can you tell the difference between the output of `lapply()` and `sapply()`? The former returns a list, while the latter returns a vector that is a simplified version of this list. Notice that this time, unlike in the cities example of the instructional video, the vector is not named.*  


####sapply with your own function  
Like `lapply()`, `sapply()` allows you to use self-defined functions and apply them over a vector or a list:  

>`sapply(X, FUN, ...)`  

Here, FUN can be one of R's built-in functions, but it can also be a function you wrote. This self-written function can be defined before hand, or can be inserted directly as an anonymous function.  

```{r}
# temp is already defined in the workspace
temp

# Finish function definition of extremes_avg
extremes_avg <- function(x) {
  ( min(x) + max(x) ) / 2
}

# Apply extremes_avg() over temp using sapply()
sapply(temp, extremes_avg)

# Apply extremes_avg() over temp using lapply()
lapply(temp, extremes_avg)
```

*Of course, you could have solved this exercise using an anonymous function, but this would require you to use the code inside the definition of `extremes_avg()` twice. Duplicating code should be avoided as much as possible!*  


####sapply with function returning vector  

In the previous exercises, you've seen how `sapply()` simplifies the list that `lapply()` would return by turning it into a vector. But what if the function you're applying over a list or a vector returns a vector of length greater than 1?    

```{r}
# temp is already available in the workspace
temp

# Create a function that returns min and max of a vector: extremes
extremes <- function(x) {
  c(min = min(x), max = max(x))
}
# Apply extremes() over temp with sapply()
sapply(temp, extremes)

# Apply extremes() over temp with lapply()
lapply(temp, extremes)
```


####sapply can't simplify, now what?  

It seems like we've hit the jackpot with `sapply()`. On all of the examples so far, `sapply()` was able to nicely simplify the rather bulky output of `lapply()`. But, as with life, there are things you can't simplify. How does `sapply()` react?  

We already created a function, `below_zero()`, that takes a vector of numerical values and returns a vector that only contains the values that are strictly below zero.  

```{r}
# temp is already prepared for you in the workspace
temp

# Definition of below_zero()
below_zero <- function(x) {
  return(x[x < 0])
}
# Apply below_zero over temp using sapply(): freezing_s
freezing_s <- sapply(temp, below_zero)

# Apply below_zero over temp using lapply(): freezing_l
freezing_l <- lapply(temp, below_zero)

# Are freezing_s and freezing_l identical?
identical(freezing_s, freezing_l)
```

*Given that the length of the output of below_zero() changes for different input vectors, sapply() is not able to nicely convert the output of lapply() to a nicely formatted matrix. Instead, the output values of sapply() and lapply() are exactly the same, as shown by the TRUE output of identical()*  


####sapply with functions that return NULL  

You already have some apply tricks under your sleeve, but you're surely hungry for some more, aren't you? In this exercise, you'll see how `sapply()` reacts when it is used to apply a function that returns `NULL` over a *vector or a list*.  

A function `print_info()`, that takes a vector and prints the average of this vector, has already been created for you. It uses the `cat()` function.  

```{r}
# Definition of print_info()
print_info <- function(x) {
  cat("The average temperature is", mean(x), "\n")
}

# Apply print_info() over temp using sapply()
sapply(temp, print_info)

# Apply print_info() over temp using lapply()
lapply(temp, print_info)
```

*Notice here that, quite surprisingly, `sapply()` does not simplify the list of `NULL's`. That's because the 'vector-version' of a list of `NULL's` would simply be a `NULL`, which is no longer a vector with the same length as the input.*  


##vapply  

####Use vapply  

Before you get your hands dirty with the third and last apply function that you'll learn about in this intermediate R course, let's take a look at its syntax. The function is called `vapply()`, and it has the following syntax:  

>`vapply(X, FUN, FUN.VALUE, ..., USE.NAMES = TRUE)`  

Over the elements inside X, the function `FUN` is applied. The `FUN.VALUE` argument expects a template for the return argument of this function `FUN`. `USE.NAMES` is TRUE by default; in this case `vapply()` tries to generate a named array, if possible.  

For the next set of exercises, you'll be working on the `temp` list again, that contains 7 numerical vectors of length 5. We also coded a function `basics()` that takes a vector, and returns a named vector of length 3, containing the minimum, mean and maximum value of the vector respectively.  

```{r}
# temp is already available in the workspace
temp

# Definition of basics()
basics <- function(x) {
  c(min = min(x), mean = mean(x), max = max(x))
}

# Apply basics() over temp using vapply()
vapply(temp, basics, numeric(3))
```

*Notice how, just as with `sapply()`, `vapply()` neatly transfers the names that you specify in the `basics()` function to the row names of the matrix that it returns.*  


####Use vapply (2)  

So far you've seen that `vapply()` mimics the behavior of `sapply()` if everything goes according to plan. But what if it doesn't?  

In the video, Filip showed you that there are cases where the structure of the output of the function you want to apply, `FUN`, does not correspond to the template you specify in `FUN.VALUE`. In that case, `vapply()` will throw an error that informs you about the misalignment between expected and actual output.  

```{r}
# temp is already available in the workspace
temp

# Definition of the basics() function
basics <- function(x) {
  c(min = min(x), mean = mean(x), median = median(x), max = max(x))
}

# Fix the error- change 3 to 4 to make it work:
vapply(temp, basics, numeric(4))
```


####From sapply to vapply  

As highlighted before, `vapply()` can be considered a more robust version of `sapply()`, because you explicitly restrict the output of the function you want to apply. Converting your `sapply()` expressions in your own R scripts to `vapply()` expressions is therefore a good practice (and also a breeze!).  

```{r}
# temp is already defined in the workspace
temp

# Convert to vapply() expression
# sapply(temp, max)
vapply(temp, max, numeric(1))

# Convert to vapply() expression
# sapply(temp, function(x, y) { mean(x) > y }, y = 5)
vapply(temp, function(x, y) { mean(x) > y }, y = 5, logical(1))
```




