---
title: "Cleaning Data in R - Preparing data for analysis"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    css: [default, extra.css]
    includes:
      after_body: ../assets/html/footer.html
---


```{r, echo=FALSE, results='hide', message = FALSE}
# Load the knitr and kableExtra packages
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
# Load the gapminder package
library(gapminder)
# Load the dpylr package
library(dplyr)
# Load the ggplot2 package as well
library(ggplot2)
theme_set(theme_bw())  # pre-set the bw theme.
```

##Preparing data

###Type conversion  

####Types of variables in R  

As in other programming languages, R is capable of storing data in many different formats, most of which you've probably seen by now.  

Loosely speaking, the `class()` function tells you what type of object you're working with. (There are subtle differences between the `class`, `type`, and `mode` of an object, but these distinctions are beyond the scope of this course.)  

```{r}
# Make this evaluate to "character"
class("TRUE")

# Make this evaluate to "numeric"
class(8484.00)

# Make this evaluate to "integer"
class(99L)

# Make this evaluate to "factor"
class(as.factor("factor"))

# Make this evaluate to "logical"
class(FALSE)
```


####Common type conversions  

It is often necessary to change, or coerce, the way that variables in a dataset are stored. This could be because of the way they were read into R (with read.csv(), for example) or perhaps the function you are using to analyze the data requires variables to be coded a certain way.  

Only certain coercions are allowed, but the rules for what works are generally pretty intuitive. For example, trying to convert a character string to a number gives an `error`: `as.numeric("some text")`.  

There are a few less intuitive results. For example, under the hood, the logical values `TRUE` and `FALSE` are coded as `1` and `0`, respectively. Therefore, `as.logical(1)` returns `TRUE` and `as.numeric(TRUE)` returns `1`.  

```{r, message = FALSE}
# Read students data
library(readr)
students <- read_csv("../xDatasets/students_with_dates.csv")

# Preview students with str()
str(students, give.attr = FALSE)

# Coerce Grades to character
students$Grades <- as.character(students$Grades)

# Coerce Medu to factor
students$Medu <- as.factor(students$Medu)

# Coerce Fedu to factor
students$Fedu <- as.factor(students$Fedu)
    
# Look at students once more with str()
str(students, give.attr = FALSE)
```


####Working with dates  

Dates can be a challenge to work with in any programming language, but thanks to the `lubridate` package, working with dates in R isn't so bad. Since this course is about cleaning data, we only cover the most basic functions from `lubridate` to help us standardize the format of dates and times in our data.  

These functions combine the letters `y`, `m`, `d`, `h`, `m`, `s`, which stand for year, month, day, hour, minute, and second, respectively. The order of the letters in the function should match the order of the date/time you are attempting to read in, although not all combinations are valid. Notice that the functions are "smart" in that they are capable of parsing multiple formats.  

```{r, eval = FALSE}
install.packages("lubridate")
```

```{r, message = FALSE}
# Read students data
library(readr)
students2 <- read_csv("../xDatasets/students_with_dates.csv")

# Preview students2 with str()
#str(students2)

# Load the lubridate package
library(lubridate)

# Parse as date
dmy("17 Sep 2015")

# Parse as date and time (with no seconds!)
mdy_hm("July 15, 2012 12:56")

# Coerce dob to a date (with no time)
students2$dob <- ymd(students2$dob)

# Coerce nurse_visit to a date and time
students2$nurse_visit <- ymd_hms(students2$nurse_visit)
    
# Look at students2 once more with str()
str(students2, give.attr = FALSE, vec.len = 8)
```


###String manipulation  

```{r, eval = FALSE}
install.packages("stringr")
```

####Trimming and padding strings  

One common issue that comes up when cleaning data is the need to remove leading and/or trailing white space. The `str_trim()` function from `stringr` makes it easy to do this while leaving intact the part of the string that you actually want.  

> `str_trim("  this is a test     ")`  
> `[1] "this is a test"`  

A similar issue is when you need to `pad strings` to make them a certain number of `characters wide`. One example is if you had a bunch of employee ID numbers, some of which begin with one or more zeros. When reading these data in, you find that the leading zeros have been dropped somewhere along the way (probably because the variable was thought to be numeric and in that case, leading zeros would be unnecessary.)  

> `str_pad("24493", width = 7, side = "left", pad = "0")`  
> `[1] "0024493"`  

```{r}
# Load the stringr package
library(stringr)

# Trim all leading and trailing whitespace
str_trim(c("   Filip ", "Nick  ", " Jonathan"))

# Pad these strings with leading zeros
str_pad(c("23485W", "8823453Q", "994Z"), width = 9, side = "left", pad = "0")
```

*Examples like this are certainly handy in R. For example, the `str_pad()` function is useful when importing a dataset with US zip codes. Occasionally R will drop the leading 0 in a zipcode, thinking it's numeric.*  


####Upper and lower case  

In addition to trimming and padding strings, you may need to adjust their case from time to time. Making strings uppercase or lowercase is very straightforward in (base) R thanks to `toupper()` and `tolower()`. Each function takes exactly one argument: the character string (or vector/column of strings) to be converted to the desired case.  


```{r}
# state abbreviations 
states <- c("al", "ak", "az", "ar", "ca", "co", "ct", "de", "fl", "ga", "hi", "id", "il", "in", "ia", "ks", "ky", "la", "me", "md", "ma", "mi", "mn", "ms", "mo", "mt", "ne", "nv", "nh", "nj", "nm", "ny", "nc", "nd", "oh", "ok", "or", "pa", "ri", "sc", "sd", "tn", "tx", "ut", "vt", "va", "wa", "wv", "wi", "wy")

# Make states all uppercase and save result to states_upper
states_upper <- toupper(states)

# Make states_upper all lowercase again
tolower(states_upper)
```


####Finding and replacing strings  

The `stringr` package provides two functions that are very useful for `finding` and/or `replacing` patterns in strings: `str_detect()` and `str_replace()`.  

Like all functions in `stringr`, the first argument of each is the string of interest. The second argument of each is the pattern of interest. In the case of `str_detect()`, this is the pattern we are searching for. In the case of `str_replace()`, this is the pattern we want to replace. Finally, `str_replace()` has a third argument, which is the string to replace with.  

> `str_detect(c("banana", "kiwi"), "a")`  
> `[1]  TRUE FALSE`  

> `str_replace(c("banana", "kiwi"), "a", "o")`  
> `"bonana" "kiwi"`  

The data.frame `students2` is already available for you in the workspace. `stringr` is already loaded. `students3` is a copy of it for you to work on so you can always start from scratch if you happen to make a mistake.  

```{r}
# Copy of students2: students3
students3 <- students2

# Look at the head of students3
students3 %>%
  head() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

# Detect all dates of birth (dob) in 1997, print 10 first results
str_detect(students3$dob, "1997")[1:10]

# In the sex column, replace "F" with "Female" ...
students3$sex <- str_replace(students3$sex, "F", "Female") 

# ... and "M" with "Male"
students3$sex <- str_replace(students3$sex, "M", "Male") 

# View the tail of students3
students3 %>%
  tail(8) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


###Missing and special values  

####Finding missing values  

As you've seen, missing values in R should be represented by `NA`, but unfortunately you will not always be so lucky. Before you can deal with missing values, you have to find them in the data.  

If missing values are properly coded as `NA`, the `is.na()` function will help you find them. Otherwise, if your dataset is too big to just look at the whole thing, you may need to try searching for some of the usual suspects like `""`, `"#N/A"`, etc. You can also use the `summary()` and `table()` functions to turn up unexpected values in your data.  

In this exercise, we've created a simple dataset called `social_df` that has 3 pieces of information for each of four friends:  

> Name  
Number of friends on a popular social media platform  
Current "status" on the platform  

```{r}
# Create small Social data frame
name <- c("Sarah", "Tom", "David", "Alice")
n_friends <- c(244, NA, 145, 43)
status <- c("Going out!", "", "Movie night...", "")
social_df <- data.frame(name, n_friends, status)

# Call is.na() on the full social_df to spot all NAs
is.na(social_df)

# Use the any() function to ask whether there are any NAs in the data
any(is.na(social_df))

# View a summary() of the dataset
summary(social_df)

# Call table() on the status column
table(social_df$status)
```

*Scanning your dataset for `NA` values is essential before learning how to remedy missing data problems.*  


####Dealing with missing values  

Missing values can be a rather complex subject, but here we'll only look at the simple case where you are simply interested in normalizing and/or removing all missing values from your data. For more information on why this is not always the best strategy, search online for "missing not at random."   

Looking at the `social_df` dataset again, we asked around a bit and figured out what's causing the missing values that you saw in the last exercise. Tom doesn't have a social media account on this particular platform, which explains why his number of friends and current status are missing (although coded in two different ways). Alice is on the platform, but is a passive user and never sets her status, hence the reason it's missing for her.  

The `stringr` package is preloaded.  

```{r}
# Replace all empty strings in status with NA
social_df$status[social_df$status == ""] <- NA

# Print social_df to the console
social_df

# Use complete.cases() to see which rows have no missing values
complete.cases(social_df)

# Use na.omit() to remove all rows with any missing values
na.omit(social_df)
```

*Often times in data analyses, you'll want to get a feel for how many complete observations you have. This can be helpful in determining how you handle observations with missing data points.*  


###Outliers and obvious errors  

```{r}
# Simulate some data with three outliers
set.seed(10)
x <- c(rnorm(30, mean = 15, sd = 5), -5, 28, 35)

# View boxplot
boxplot(x, horizontal = TRUE)
```


####Dealing with outliers and obvious errors  

When dealing with strange values in your data, you often must decide whether they are just extreme or actually erroneous. Extreme values show up all over the place, but you, the data analyst, must figure out when they are plausible and when they are not.  

We have loaded a dataset called `students3`, which is another slight variation of the original `students` dataset. Two variables appear to have suspicious values: `age` and `absences`. Let's explore these values further.  

```{r, message = FALSE}
# Read students data
students3 <- read_csv("../xDatasets/students_with_dates.csv")

# Simulate AGE and ABSCENCES variables
students3$age <- sample(15:40, size = nrow(students3), replace = TRUE)

# Look at a summary() of students3
sum_students3 <- as.data.frame(do.call(cbind, lapply(students3, summary)))

sum_students3[,-1] %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

# View a histogram of the age variable
hist(students3$age)

# View a histogram of the absences variable
hist(students3$absences)

# View a histogram of absences, but force zeros to be bucketed to the right of zero
hist(students3$absences, right = FALSE)
```

*As you can see, a simple histogram, displaying the distribution of a variable's values across all the observations can be key to identifying potential outliers as early as possible.*  


####Another look at strange values  

Another useful way of looking at strange values is with boxplots. Simply put, boxplots draw a box around the middle 50% of values for a given variable, with a bolded horizontal line drawn at the median. Values that fall far from the bulk of the data points (i.e. outliers) are denoted by open circles. (If you're curious about the exact formula for determining what is "far", check out `?hist.`)

```{r}
# View a boxplot of age
boxplot(students3$age)

# View a boxplot of absences
boxplot(students3$absences)

```

