---
title: "Cleaning Data in R - It's your turn"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    css: [default, extra.css]
    includes:
      after_body: ../assets/html/footer.html
---


```{r, echo=FALSE, results='hide', message = FALSE}
# Load the knitr and kableExtra packages
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
# Load the gapminder package
library(gapminder)
# Load the dpylr package
library(dplyr)
# Load the ggplot2 package as well
library(ggplot2)
theme_set(theme_bw())  # pre-set the bw theme.
```

###Get a feel for the data  

Before diving into our data cleaning routine, we must first understand the basic structure of the data. This involves looking at things like the `class()` of the data object to make sure it's what we expect (generally a data.frame) in addition to checking its dimensions with `dim()` and the column names with `names()`.  

```{r, message = FALSE}
# Read weather data
library(readr)
weather <- readRDS("../xDatasets/weather.rds")
```

```{r}
# Verify that weather is a data.frame
class(weather)

# Check the dimensions
dim(weather)

# View the column names
names(weather)
```

*We've confirmed that the object is a data frame with 286 rows and 35 columns. We'll see what the columns represent in the upcoming exercises.*  


###Summarize the data  

Next up is to look at some summaries of the data. This is where functions like `str()`, `glimpse()` from dplyr, and `summary()` come in handy.  

```{r}
# View the structure of the data
str(weather, give.attr = FALSE)

# Load dplyr package
library(dplyr)

# Look at the structure using dplyr's glimpse()
glimpse(weather)

# View a summary of the data
sum_weather <- as.data.frame(do.call(cbind, lapply(weather, summary)))

sum_weather[,-1] %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


###Take a closer look  

After understanding the structure of the data and looking at some brief summaries, it often helps to preview the actual data. The functions `head()` and `tail()` allow you to view the `top` and `bottom` rows of the data, respectively. Recall you'll be shown 6 rows by default, but you can alter this behavior with a second argument to the function.  

```{r}
# View first 15 rows
weather %>% 
  head(15) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

# View the last 10 rows
weather %>% 
  tail(10) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


###Column names are values  

The weather dataset suffers from one of the five most common symptoms of messy data: `column names are values`. In particular, the column names `X1-X31` represent days of the month, which should really be values of a new variable called day.  

The tidyr package provides the `gather()` function for exactly this scenario. 

> `gather(df, time, val, t1:t3)`  

*Notice that `gather()` allows you to select multiple columns to be gathered by using the : operator.*  

```{r}
# Load the tidyr package
library(tidyr)

# Gather the columns
weather2 <- gather(weather, day, value, X1:X31, na.rm = TRUE)

# View the head
weather2 %>%
  head() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```



###Values are variable names  

Our data suffer from a `second common symptom of messy data`: values are variable names. Specifically, values in the measure column should be variables (i.e. column names) in our dataset.

The `spread()` function from `tidyr` is designed to help with this. 

> `spread(df2, time, val)`  

*Note how the values of the time column now become column names. The tidyr package is already loaded.*  

```{r}
# First remove column of row names
without_x <- weather2[, -1]

# Spread the data
weather3 <- spread(without_x, measure, value)

# View the head
weather3 %>%
  head() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "center", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```

*The dataset is looking better already!*  


###Clean up dates  

Now that the weather dataset adheres to tidy data principles, the next step is to prepare it for analysis. We'll start by `combining` the `year`, `month`, and `day` columns and recoding the resulting character column as a `date`. We can use a combination of base R, `stringr`, and `lubridate` to accomplish this task.  

`tidyr` and `dplyr` are already loaded.  

```{r, message = FALSE}
# Load the stringr and lubridate packages
library(stringr)
library(lubridate)

# Remove X's from day column
weather3$day <- str_replace(weather3$day, "X", "")

# Unite the year, month, and day columns
weather4 <- unite(weather3, date, year, month, day, sep = "-")

# Convert date column to proper date format using lubridates's ymd()
weather4$date <- as.Date(weather4$date)

# Rearrange columns using dplyr's select()
weather5 <- select(weather4, date, Events, CloudCover:WindDirDegrees)

# View the head of weather5
weather5 %>%
  head() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


###A closer look at column types  

It's important for analysis that variables are coded appropriately. This is not yet the case with our weather data. Recall that functions such as `as.numeric()` and `as.character()` can be used to coerce variables into different types.  

It's important to keep in mind that coercions are `not always successful`, particularly if there's some data in a column that you don't expect. For example, the following will cause problems:  

> `as.numeric(c(4, 6.44, "some string", 222))`  

If you run the code above in the console, you'll get a warning message saying that R introduced an `NA` in the process of coercing to numeric. This is because it doesn't know how to make a number out of a string ("some string").  

```{r}
# View the structure of weather5
str(weather5, give.attr = FALSE)

# Examine the first 20 rows of weather5. Are most of the characters numeric?
weather5 %>%
  select(date, Events, CloudCover, Max.Humidity, PrecipitationIn) %>%
  head(20) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

# See what happens if we try to convert PrecipitationIn to numeric
as.numeric(weather5$PrecipitationIn)[1:10]
```

*Scroll the output, notice the warning message. Go back to the results of the head command if need be. What values in PrecipitationIn would become `NA` if coerced to numbers? Why would they be in the dataset to begin with?*  


###Column type conversions  

As you saw in the last exercise, `"T"` was used to denote a `trace amount` (i.e. too small to be accurately measured) of precipitation in the PrecipitationIn column. In order to coerce this column to numeric, you'll need to deal with this somehow. To keep things simple, we will just replace `"T"` with `zero`, as a `string ("0")`.  

The `dplyr` and `stringr` packages are already loaded!  

```{r, message = FALSE}
# Replace "T" with "0" (T = trace)
weather5$PrecipitationIn <- str_replace(weather5$PrecipitationIn, "T", "0")

# Convert characters to numerics
weather6 <- mutate_at(weather5, vars(CloudCover:WindDirDegrees), funs(as.numeric))

# Look at result
str(weather6, give.attr = FALSE)
```

*It looks like our data are finally in the correct formats and organized in a logical manner! Now that our data are in the right form, we can begin the analysis.*  


###Find missing values  

Before dealing with missing values in the data, it's important to find them and figure out why they exist in the first place. If your dataset is too big to look at all at once, like it is here, remember you can use `sum()` and `is.na()` to quickly size up the situation by counting the number of `NA` values.  

The `summary()` function may also come in handy for identifying which variables contain the missing values. Finally, the `which()` function is useful for `locating` the `missing values within a particular column`.  

```{r, message = FALSE}
# Count missing values
sum(is.na(weather6))

# Find missing values
sum_weather6 <- as.data.frame(do.call(cbind, lapply(weather6, summary)))

sum_weather6 [,-1] %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

# Find indices of NAs in Max.Gust.SpeedMPH
ind <- which(is.na(weather6$Max.Gust.SpeedMPH))

# Look at the full rows for records missing Max.Gust.SpeedMPH
weather6[ind, ] %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```

*In this situation it's unclear why these values are missing and there doesn't appear to be any obvious pattern to their missingness, so we'll leave them alone for now.*  


###An obvious error  

Besides missing values, we want to know if there are values in the data that are too extreme or bizarre to be plausible. A great way to start the search for these values is with `summary()`.  

Once implausible values are identified, they must be dealt with in an intelligent and informed way. Sometimes the best way forward is obvious and other times it may require some research and/or discussions with the original collectors of the data.  

```{r, message = FALSE}
# Review distributions for all variables (see above)
#summary(weather6)

# Find row with Max.Humidity of 1000
ind <- which(weather6$Max.Humidity == 1000)

# Look at the data for that day
weather6[ind, ] %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

# Change 1000 to 100
weather6$Max.Humidity[ind] <- 100
```

*Once you find obvious errors, it's not too hard to fix them if you know which values they should take.*  


###Another obvious error  

You've discovered and repaired one obvious error in the data, but it appears that there's another. Sometimes you get lucky and can infer the correct or intended value from the other data. For example, if you know the minimum and maximum values of a particular metric on a given day...  

```{r}
# Look at summary of Mean.VisibilityMiles
summary(weather6$Mean.VisibilityMiles)

# Get index of row with -1 value
ind <- which(weather6$Mean.VisibilityMiles == -1)

# Look at full row
weather6[ind,] %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")

# Set Mean.VisibilityMiles to the appropriate value
weather6$Mean.VisibilityMiles[ind] <- 10
```


###Check other extreme values  

In addition to dealing with obvious errors in the data, we want to see if there are other extreme values. In addition to the trusty `summary()` function, `hist()` is useful for quickly getting a feel for how different variables are distributed.  

```{r}
# Look at histogram for MeanDew.PointF
hist(weather6$MeanDew.PointF)

# Look at histogram for Min.TemperatureF
hist(weather6$Min.TemperatureF)

# Compare to histogram for Mean.TemperatureF
hist(weather6$Mean.TemperatureF)
```


###Finishing touches  

Before officially calling our weather data clean, we want to put a couple of finishing touches on the data. These are a bit more subjective and may not be necessary for analysis, but they will make the data easier for others to interpret, which is generally a good thing.  

There are a number of stylistic conventions in the R language. Depending on who you ask, these conventions may vary. Because the period `(.)` has special meaning in certain situations, we generally recommend using underscores `(_)` to separate words in variable names. We also `prefer` all `lowercase letters` so that no one has to remember which letters are uppercase or lowercase.  

Finally, the events column (renamed to be all lowercase in the first instruction) contains an empty string `("")` for any day on which there was no significant weather event such as rain, fog, a thunderstorm, etc. However, if it's the first time you're seeing these data, it may not be obvious that this is the case, so it's best for us to be explicit and replace the empty strings with something more meaningful.  

```{r}
new_colnames <- c("date", "events", "cloud_cover", "max_dew_point_f",         "max_gust_speed_mph", "max_humidity", "max_sea_level_pressure_in", "max_temperature_f", "max_visibility_miles", "max_wind_speed_mph",       "mean_humidity", "mean_sea_level_pressure_in", "mean_temperature_f",        "mean_visibility_miles", "mean_wind_speed_mph", "mean_dew_point_f",           
"min_dew_point_f", "min_humidity", "min_sea_level_pressure_in", "min_temperature_f", "min_visibility_miles", "precipitation_in", "wind_dir_degrees")

# Clean up column names
names(weather6) <- new_colnames

# Replace empty cells in events column
weather6$events[weather6$events == ""] <- "None"
    
# Print the first 6 rows of weather6
weather6 %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689") %>%
  scroll_box(width = "100%", height = "500px")
```

*Your data are now tidy and in an easy format for others to examine!*  

```{r, message = FALSE}
# Save our tidy dataframe to csv file
write.csv(weather6,'../xDatasets/weather_clean.csv')
```

