---
title: "Import data with R - Importing data from databases (Part 1)"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    css: [default, extra.css]
    includes:
      after_body: ../assets/html/footer.html
---

```{r, echo=FALSE, results='hide', message = FALSE}
# Load the knitr and kableExtra packages
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
# Load the gapminder package
library(gapminder)
# Load the dpylr package
library(dplyr)
# Load the ggplot2 package as well
library(ggplot2)
theme_set(theme_bw())  # pre-set the bw theme.
```

##RMySQL package  

```{r, echo=FALSE, results='hide', message = FALSE, eval=FALSE}
install.packages("RMySQL")
```


####Establish a connection  

The first step to import data from a SQL database is creating a connection to it. As Filip explained, you need different packages depending on the database you want to connect to. All of these packages do this in a uniform way, as specified in the DBI package.  

`dbConnect()` creates a connection between your R session and a SQL database. The first argument has to be a DBIdriver object, that specifies how connections are made and how data is mapped between R and the database. Specifically for MySQL databases, you can build such a driver with `RMySQL::MySQL()`.  

If the MySQL database is a remote database hosted on a server, you'll also have to specify the following arguments in dbConnect(): `dbname, host, port, user and password`. Most of these details have already been provided.  

```{r}
# Load the DBI package
library(DBI)

# Edit dbConnect() call
con <- dbConnect(RMySQL::MySQL(), 
                 dbname = "tweater", 
                 host = "courses.csrrinzqubik.us-east-1.rds.amazonaws.com", 
                 port = 3306,
                 user = "student",
                 password = "datacamp")
```


####List the database tables  

After you've successfully connected to a remote MySQL database, the next step is to see what tables the database contains. You can do this with the `dbListTables()` function. As you might remember, this function requires the connection object as an input, and outputs a character vector with the table names.  

```{r}
# Load the DBI package
library(DBI)

# Connect to the MySQL database: con
con <- dbConnect(RMySQL::MySQL(), 
                 dbname = "tweater", 
                 host = "courses.csrrinzqubik.us-east-1.rds.amazonaws.com", 
                 port = 3306,
                 user = "student",
                 password = "datacamp")

# Build a vector of table names: tables
tables <- dbListTables(con)


# Display structure of tables
str(tables)
```

* `dbListTables()` can be very useful to get a first idea about the contents of your database. Can you guess what kind of information this database contains?*  


####Import a table (users)  

As you might have guessed by now, the database contains data on a more tasty version of Twitter, namely Tweater. Users can post tweats with short recipes for delicious snacks. People can comment on these tweats. There are three tables: users, tweats, and comments that have relations among them. Which ones, you ask? You'll discover in a moment!  

Let's start by importing the data on the users into your R session. You do this with the `dbReadTable()` function. Simply pass it the connection object (con), followed by the name of the table you want to import. The resulting object is a standard R data frame.  

```{r}
# Load the DBI package
library(DBI)

# Connect to the MySQL database: con
con <- dbConnect(RMySQL::MySQL(), 
                 dbname = "tweater", 
                 host = "courses.csrrinzqubik.us-east-1.rds.amazonaws.com", 
                 port = 3306,
                 user = "student",
                 password = "datacamp")

# Import the users table from tweater: users
users <- dbReadTable(con, "users")

# Print users
users %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


####Import all tables  

Next to the users, we're also interested in the tweats and comments tables. However, separate `dbReadTable()` calls for each and every one of the tables in your database would mean a lot of code duplication. Remember about the `lapply()` function? You can use it again here! A connection is already coded for you, as well as a vector `table_names`, containing the names of all the tables in the database.  

```{r, eval = FALSE}
# Load the DBI package
library(DBI)

# Connect to the MySQL database: con
con <- dbConnect(RMySQL::MySQL(), 
                 dbname = "tweater", 
                 host = "courses.csrrinzqubik.us-east-1.rds.amazonaws.com", 
                 port = 3306,
                 user = "student",
                 password = "datacamp")

# Get table names
table_names <- dbListTables(con)

# Import all tables
tables <- lapply(table_names, dbReadTable, conn = con)

# Print out tables
tables %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```

