---
title: "TidyVerse - Data wrangling"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    css: ["default", "extra.css"]
    includes:
      after_body: footer.html
---


##Data wrangling 

####Loading the gapminder and dplyr packages  

Before you can work with the gapminder dataset, you'll need to load two R packages that contain the tools for working with it, then display the gapminder dataset so that you can see what it contains.  

```{r, echo=TRUE, results='hide', message = FALSE}
# Load the knitr and kableExtra packages
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
# Load the gapminder package
library(gapminder)
# Load the dpylr package
library(dplyr)
```

```{r}
# Look at the gapminder dataset
gapminder %>%
  kable(caption = "Gapminder dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689") %>%
  scroll_box(width = "100%", height = "300px")
```


####Filtering for one year  

The `filter` verb extracts particular observations based on a condition. In this exercise you'll filter for observations from a particular year.  

```{r}
# Filter the gapminder dataset for the year 1957
gapminder %>%
  filter(year == 1957) %>% 
  head %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


####Filtering for one country and one year  

You can also use the `filter()` verb to set two conditions, which could retrieve a single observation.  

Just like in the last exercise, you can do this in two lines of code, starting with `gapminder %>%` and having the `filter()` on the second line. Keeping one verb on each line helps keep the code readable. Note that each time, you'll put the `pipe %>%` at the end of the first line (like `gapminder %>%`); putting the pipe at the beginning of the second line will throw an error.  

```{r}
# Filter for China in 2002
gapminder %>%
  filter(country == "China", year== 2002) %>% 
  kable(caption = "GDB of China in 2002") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


#### Arranging observations by life expectancy  

You use `arrange()` to sort observations in ascending or descending order of a particular variable. In this case, you'll sort the dataset based on the `lifeExp` variable.  

```{r}
# Sort in descending order of lifeExp
gapminder %>%
  select("country", "lifeExp") %>% 
  arrange(desc(lifeExp)) %>% 
  head(5) %>%
  kable(caption = "5 top countries with the highest life expectancy") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")


# Sort in ascending order of lifeExp
gapminder %>%
  select("country", "lifeExp") %>% 
  arrange(lifeExp) %>% 
  head(5) %>%
  kable(caption = "5 top countries with the lowest life expectancy") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


####Filtering and arranging  

You'll often need to use the pipe operator (`%>%`) to combine multiple `dplyr` verbs in a row. In this case, you'll combine a `filter()` with an `arrange()` to find the highest population countries in a particular year.  

```{r}
# Filter for the year 1957, then arrange in descending order of population
gapminder %>%
  filter(year == 1957) %>%
  arrange(desc(pop)) %>%
  head(5) %>%
  kable(caption = "Highest population countries in 1957", font_size = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left", font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689")
```


