---
title: "TidyVerse - Data visualization"
author: "Guillaume Abgrall"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    css: [default, extra.css]
    includes:
      after_body: footer.html
---

##Data visualization  

####Variable assignment  

Throughout the exercises in this chapter, you'll be visualizing a subset of the gapminder data from the year 1952. First, you'll have to load the `ggplot2` package, and create a `gapminder_1952` dataset to visualize.  

```{r, echo=TRUE, results='hide', message = FALSE}
# Load the knitr and kableExtra packages
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
# Load the gapminder package
library(gapminder)
# Load the dpylr package
library(dplyr)
# Load the ggplot2 package as well
library(ggplot2)
theme_set(theme_bw())  # pre-set the bw theme.
```

```{r}
# Create gapminder_1952
gapminder_1952 <- gapminder %>%
    filter(year == 1952)
```

```{r}
# Look at the gapminder_1952 dataset
gapminder_1952 %>%
  kable(caption = "Gapminder from 1952") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = T, position = "left", , font_size = 11) %>%
  row_spec(0, bold = T, color = "white", background = "#3f7689") %>%
  scroll_box(width = "100%", height = "300px")
```


####Comparing population and GDP per capita 

In the video you learned to create a scatter plot with GDP per capita on the x-axis and life expectancy on the y-axis (the code for that graph is shown here). When you're exploring data visually, you'll often need to try different combinations of variables and aesthetics.  

```{r}
ggplot(gapminder_1952, aes(x = pop, y = gdpPercap)) +
  geom_point() + 
  geom_smooth(method="loess", se=F) +
  labs(subtitle="GDP by capita by population", 
       y="GDP per capita", 
       x="Population", 
       title="Scatterplot", 
       caption = "")
```

*Each point represents a country: can you guess which country any of the points are?*  


####Comparing population and life expectancy  

In this exercise, you'll use `ggplot2` to create a scatter plot from scratch, to compare each country's population with its life expectancy in the year 1952.  

```{r}
# Create a scatter plot with pop on the x-axis and lifeExp on the y-axis
ggplot(gapminder_1952, aes(x = pop, y = lifeExp)) +
  geom_point()+
  geom_smooth(method="loess", se=F) +
  labs(subtitle="Country's population with its life expectancy in the year 1952", 
       y="Life Expectancy", 
       x="Population", 
       title="Scatterplot", 
       caption = "")
```

*You might notice the points are crowded towards the left side of the plot, making them hard to distinguish.*  


####Putting the x-axis on a log scale  

You previously created a scatter plot with population on the x-axis and life expectancy on the y-axis. Since population is spread over several orders of magnitude, with some countries having a much higher population than others, it's a good idea to put the x-axis on a log scale.  


```{r}
# Change this plot to put the x-axis on a log scale
ggplot(gapminder_1952, aes(x = pop, y = lifeExp)) +
  geom_point()+
  scale_x_log10() +
  geom_smooth(method="loess", se=F) +
  labs(subtitle="Country's population (passed into log scale) with its life expectancy in the year 1952", 
       y="Life Expectancy", 
       x="Population", 
       title="Scatterplot", 
       caption = "")
```

*Notice the points are more spread out on the x-axis. This makes it easy to see that there isn't a correlation between population and life expectancy.*


####Putting the x- and y- axes on a log scale  

Suppose you want to create a scatter plot with population on the x-axis and GDP per capita on the y-axis. Both population and GDP per-capita are better represented with log scales, since they vary over many orders of magnitude.  

```{r}
# Scatter plot comparing pop and gdpPercap, with both axes on a log scale
ggplot(gapminder_1952, aes(x = pop, y = gdpPercap)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() + 
  geom_smooth(method="loess", se=F) +
  labs(subtitle="Country's population (log scale) with GDP by capita (log scale) in the year 1952", 
       y="GDP by capita", 
       x="Population", 
       title="Scatterplot", 
       caption = "")
```

*Notice that the y-axis goes from 1e3 (1000) to 1e4 (10,000) to 1e5 (100,000) in equal increments.*  

####Adding color to a scatter plot  

In this lesson you learned how to use the color aesthetic (`color` and `pop`), which can be used to show which continent each point in a scatter plot represents.  

```{r}
# Scatter plot comparing pop and lifeExp, with color representing continent
ggplot(gapminder_1952, aes(x = pop, y = lifeExp, color = continent))+
  geom_point() +
  scale_x_log10() +
  labs(subtitle="Country's population (log scale) with Life expectancy in the year 1952", 
       y="Life expectancy", 
       x="Population", 
       title="Scatterplot colored by continent", 
       caption = "")
```


####Adding size and color to a plot  

In the last exercise, you created a scatter plot communicating information about each country's population, life expectancy, and continent. Now you'll use the `size` of the points to communicate even more.  

```{r}
# Add the size aesthetic to represent a country's gdpPercap
ggplot(gapminder_1952, aes(x = pop, y = lifeExp, color = continent, size = gdpPercap)) +
  geom_point() +
  scale_x_log10() +
  labs(subtitle="Country's population (log scale) with Life expectancy in the year 1952", 
       y="Life expectancy", 
       x="Population", 
       title="Scatterplot colored by continent, size by GDB by capita", 
       caption = "")
```


####Creating a subgraph for each continent  

You've learned to use `faceting` to divide a graph into subplots based on one of its variables, such as the continent.  

```{r}
# Scatter plot comparing pop and lifeExp, faceted by continent
ggplot(gapminder_1952, aes(x = pop, y =lifeExp)) +
  geom_point() +
  scale_x_log10() +
  facet_wrap(~ continent) +
  labs(subtitle="Country's population (log scale) with Life expectancy in the year 1952 by Continent", 
       y="Life expectancy", 
       x="Population", 
       title="Scatterplot of each continent", 
       caption = "")
```

*Faceting is a powerful way to understand subsets of your data separately.*  

####Faceting by year  

All of the graphs in this chapter have been visualizing statistics within one year. Now that you're able to use faceting, however, you can create a graph showing all the country-level data from 1952 to 2007, to understand how global statistics have changed over time.  

```{r}
# Scatter plot comparing gdpPercap and lifeExp, with color representing continent
# and size representing population, faceted by year
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) + 
  geom_point() +
  scale_x_log10() + 
  facet_wrap(~ year) +
  labs(subtitle="GDB per capita (log scale) with Life expectancy by Continent and size population", 
       y="Life expectancy", 
       x="GDP per Capita", 
       title="Scatterplot, every 5 years from 1952 to 2007", 
       caption = "")

```

