---
title: "1 - R basics"
author: "Guillaume Abgrall"
date: "29 Januar 2019"
output:
  html_document:
    df_print: paged
vignette: |
  %\VignetteIndexEntry{Selectorgadget} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
embed_png <- function(path, dpi = NULL) {
  meta <- attr(png::readPNG(path, native = TRUE, info = TRUE), "info")
  if (!is.null(dpi)) meta$dpi <- rep(dpi, 2)
  knitr::asis_output(paste0(
    "<img src='", path, "'",
    " width=", round(meta$dim[1] / (meta$dpi[1] / 96)),
    " height=", round(meta$dim[2] / (meta$dpi[2] / 96)),
    " />"
  ))
}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

## Calculating total winnings
Now that you have the poker and roulette winnings nicely as named vectors, you can start doing some data analytical magic.

You want to find out the following type of information:

  How much has been your overall profit or loss per day of the week?
  Have you lost money over the week in total?
  Are you winning/losing money on poker or on roulette?
  To get the answers, you have to do arithmetic calculations on vectors.

*It is important to know that if you sum two vectors in R, it takes the element-wise sum.* For example, the following three statements are completely equivalent:

```{r, echo=FALSE}
c(1, 2, 3) + c(4, 5, 6)
c(1 + 4, 2 + 5, 3 + 6)
c(5, 7, 9)
```

You can also do the calculations with variables that represent vectors:

## Including Plots

You can also embed plots, for example:

```{r, echo=FALSE}
a <- c(1, 2, 3) 
b <- c(4, 5, 6)
c <- a + b
```

## Instructions

1. Take the sum of the variables A_vector and B_vector and assign it to total_vector.
2. Inspect the result by printing out total_vector.

```{r, echo=FALSE}
A_vector <- c(1, 2, 3)
B_vector <- c(4, 5, 6)

# Take the sum of A_vector and B_vector
total_vector <- A_vector + B_vector
  
# Print out total_vector
total_vector
```

